Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-15 pecomplexity.tex
cleaning and revising

\section{Introduction}\label{sec:intro}
%\begin{it}
%Paragraph on computer performance, including citations to Todd paper
%and summary of the results that indicate that they're deterministic
%nonlinear dynamical systems.  Given that, we should be able to
%predict.  What benefits would accrue if we could do so: power mgmt,
%end world hunger [[this is my primary goal everyday :)]], etc.
%\end{it}
Things to add to introduction
\begin{enumerate}
\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.

@@

\subsection{Comparing Prediction Accuracy}
In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}. The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary. In particular, we perform $k$ 1-step predictions, after each 1-step prediction we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction. This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.}

@@

@@

%%FIGURE INTENT
%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
  \centering
         \caption{
[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly. Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD. The gray bands
    are regions where the window overlaps regimes. The window size used is
    $5,000 \times 100,000$ instructions and the word length is $4$.}\label{fig:wwpe}
  \begin{subfigure}{0.3\textwidth}
    \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
    \caption{The instructions per cycle of \svd. Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}. From left to right each change in color represents a change in regime for 6 regimes in total. }
    \label{fig:svd_ts}
  \end{subfigure}%
  %\\
  \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
    \caption{The MASE of ARIMA vs weighted permutation entropy. }
    \label{fig:svd_wwpe}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   % \caption{The MASE values for LMA against ARIMA. The dashed line is the
   % identity, delineating the traces for which either LMA or ARIMA %performed
%    better. All traces except those from {\tt SVD$_1$} lie above the line,
%    indicating that LMA is better suited prediction method for the %computer
%   traces considered.}\label{fig:lma_vs_arima}
  %\end{subfigure}
\end{figure}

@@

OUTLINE:
\begin{enumerate}
\item \cmark~Introduce MASE.
\item \cmark~WPE is a good measure of predictability.  Figure: best athlete MASE vs.
its WPE.
\item \cmark~Talk about structure analysis. Just because there is forward information
transfer, does not mean that linear predictors can get at this.
\subitem \cmark~For this show a figure of (a) ARIMA vs MASE (b) LMA vs MASE in a side by
side plot.
\item full results.  Image: MASE vs. WPE for both LMA \& ARIMA.  Points to make:
\begin{enumerate}
\item \cmark~clusters are distributed differently
\item \cmark~clusters are shaped differently---tight or not
\item \cmark~clusters move differently between LMA and ARIMA
\item finally, the diagonal line is important. If you're below it, you could do
better.
\end{enumerate}
\end{enumerate}

@@

\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"

@@

\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure). The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.

\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.

>>>>

\section{Introduction}\label{sec:intro}

%\begin{it}
%Paragraph on computer performance, including citations to Todd paper
%and summary of the results that indicate that they're deterministic
%nonlinear dynamical systems.  Given that, we should be able to
%predict.  What benefits would accrue if we could do so: power mgmt,
%end world hunger [[this is my primary goal everyday :)]], etc.
%\end{it}
Things to add to introduction
\begin{enumerate}
\item Different kinds of complexity exist in time series and this makes choosing prediction models difficult 
\subitem NOTE: RW and chaos are both complex. One is predictable and one is not.

@@

\subsection{Comparing Prediction Accuracy}
In order to analyze correctness of each prediction we split each time series into two pieces: the first 90\% referred to as the ``learning" or ``training" signal, $\{X_{i,obs}\}_{i=1}^{n}$ and the last 10\% known as the ``test" or ``correct" signal $\{c_j\}_{j=n+1}^{k+n+1}$. The learning signal is used to train an initial model (e.g., LMA or ARIMA) as described in Section \ref{sec:compModel}. The test signal is used both to assess the models forecasting accuracy and for any refitting that may be necessary. In particular, we perform $k$ 1-step predictions, after each 1-step prediction\footnote{We would like to note that this rebuilding occurs due to a problem with ARIMA models converging to a mean prediction if too long of a prediction horizon is used, this is not a handicap of either LMA or na\"ive.} we append the training signal with the next point in the correct signal $c_j$, refit the model taking into account the new system measurement and perform another prediction. This is repeated $k$ times to obtain $\{p_j\}_{j=n+1}^{k+n+1}$.

@@

THINGS TO DO
\begin{enumerate}
\item rewrite for consistency in notation. For example, some parts of the paper call word length $l$ and some call it $n$.
\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
\item justify and explain the regime split in \svd 
\end{enumerate}

@@

%%FIGURE INTENT
%This figure shows that entropy changs over time with the signal as well as why the regimes for \svd were chosen. 
%%%%%%%%%%%%%%%%%
\begin{figure}[htbp]
  \centering

  %\begin{subfigure}{0.3\textwidth}
  %  \includegraphics[width=0.3\textwidth]{figs/svdipcregimescolored.png}
  %  \caption{The instructions per cycle of \svd. Each color corresonds to the different regimes as selected by rapid shifts in WPE, as seen in Figure \ref{fig:svd_wwpe}. From left to right each change in color represents a change in regime for 6 regimes in total. }
 %   \label{fig:svd_ts}
 % \end{subfigure}%
  %\\
 % \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
  \caption{
[Joshua: I think adding the colored SVD trace to this would be good or putting it above this figure but need to figure how to line them up properly. Also we need to label that the numbers on the bottom of WPE are regimes not instructions...]]The weighted permutation entropy of one run of SVD. The gray bands
    are regions where the window overlaps regimes. The window size used is
    5,000 $\times$ 100,000 instructions and the word length is $4$.[[Joshua:why is $l=4$ and not 5,6 like the rest of the paper?]]For reference the instructions per cycle of \svd are plotted as a ghost behind this plot. Each color on the ghosted time series corresonds to the different regimes as selected by rapid shifts in WPE. From left to right each change in color represents a change in regime for 6 regimes in total.}\label{fig:wwpe}
\end{figure}

@@

%OUTLINE:
%\begin{enumerate}
%\item \cmark~Introduce MASE.
%\item \cmark~WPE is a good measure of predictability.  %Figure: best athlete MASE vs.
%its WPE.
%\item \cmark~Talk about structure analysis. Just because %there is forward information
%transfer, does not mean that linear predictors can get at %this.
%\subitem \cmark~For this show a figure of (a) ARIMA vs MASE %(b) LMA vs MASE in a side by%
%side plot.
%\item full results.  Image: MASE vs. WPE for both LMA \& %ARIMA.  Points to make:
%\begin{enumerate}
%\item \cmark~clusters are distributed differently
%\item \cmark~clusters are shaped differently---tight or not
%\item \cmark~clusters move differently between LMA and ARIMA
%\item finally, the diagonal line is important. If you're %below it, you could do
%better.
%\end{enumerate}
%\end{enumerate}

%BRAINSTORMING 
%\begin{itemize}
%\cmark\item The kind of complexity present matters, i.e., that is whether the complexity is structured or not.[[use here and mention in intro]] 
%\cmark\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
%\item Maybe plot a big chunk of \col and a big chunk of \gcc together and show that they both look complex.

%\cmark\item \gcc appears visually very complex, *and* according to WPE this complexity is unstructured. And a constant, linear and nonlinear prediction strategy all fail. We should be able to conclude that guessing random values is the best we can do as is shown by MASE [[Use in this section]]

%\cmark\item \col is also complex (can even be chaotic/ point to CHAOS paper) but the complexity is structured according to WPE and as such that complexity is usable for prediction [[use in this section]]

%\cmark\item \col brings about the point nicely that some prediction strategies cannot utilize the processes internal information transfer method. That is a nonlinear internal information transfer system cannot be predicted effectively with a linear strategy. This gives a practitioner leverage on when to give up and when to keep working. [[use in this section as bridge to next section]]

%\end{itemize}

@@

%\subitem [[Simply reminders not to be included]]WPE can quantify when a noisy real-valued time series is predictable.[[I am unsatisfied with predictable in this sentence, need a better word to say "better than random walk" or ``able to be forecast effectively" or ``has the structural capacity to transmit information in a way that the time series can be effectively forecast"

@@

%\subitem [[Simply reminders not to be included]]We will have shown that the existence whether linear or nonlinear is picked up on with WPE but this point gets at whether the prediction model can use the structure or not (linear can't use nonlinear structure). The right to left shifts in \col and some of the \svd regimes and the lack of shift in \gcc illustrate this nicely.

%\subitem [[Simply reminders not to be included]]This is the linear vs nonliear vs random we see with the right to left shifts with lower complexity time series.
